Draft Article

### 2019-11-18 article.tex
new script to parse citations

>>>>

\section{Aerial espionage}
\section{Introduction}
According to \emph{The New York Times}, the very first reconnaissance flight took place in 1911, only eight years after Kitty Hawk, when ``Lieut. Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power'' surveilling the Turkish Infantry near Tripoli.\footcite{special_cable_to_the_new_york_times_air_1911} The ability to fly over an enemy's position to gain intelligence about their movements---in a craft orders of magnitude more versatile than a balloon---was so obviously valuable that countermeasures quickly followed. Early aerial dogfighting famously took the form of pilots firing handguns at each other out of open cockpits in mid-air (emphasis on attempted), but by the time of the First World War, planes were already outfitted with synchronized machine guns and ground regiments equipped with anti-aircraft weaponry. Crucial information about military movements was going to come at a cost.

Even though these aircraft were performing vital intelligence-gathering functions, that did not make them spy planes. Aerial reconnaissance first appeared in the press in 1911, but the term ``spy plane'' appeared in the pages of \emph{The New York Times} just once before 1960.\footnote{According to my search of the archives. The instance mentioned here was in 1944, when the AP wrote that ``Germans tried to probe the secrets of the gathering Allied storm by E-boat patrol dashes [...] and by spy-plane coastal raids.''(\cite{the_associated_press_britons_1944}) In context of the article, though, it's clear that they are referring to military aerial reconnaissance, especially since it was written during the Second World War. This is a marked difference from a plane created for peacetime intelligence, which is what the term ``spy plane'' has since come to mean.} The image of a spy plane crashed into the public consciousness that year when Gary Powers crawled out of a U-2 class aircraft over Soviet Russia. The downed craft was clearly not a generic fighter plane that had flown too close to the border, and it suddenly became necessary to categorize this new type of plane, flying at previously unheard-of altitudes, that was purpose-built for clandestine photography. The United States first claimed that the lost plane was a ``weather craft.''\footcite{caruthers_soviet_1960} When that cover didn't hold up, a new story was born. The diplomatic fallout of the U-2 shootdown, which threatened to torpedo an upcoming NATO-Warsaw summit, had become the ``spy-plane case.''\footcite[This is the second time that ``spy plane'' as a term of art appeared in \emph{The New York Times}. There would be many more.]{reston_allies_1960}

I emphasize the linguistic distinction between a military plane used for reconnaissance and a ``spy plane'' because that distinction has a major impact on how the flight is perceived and what diplomatic repercussions might result when it is caught.\footnote{That is not to say there is a hard boundary between military and non-military intelligence, either in purpose or collection. The overlap between military and civilian intelligence agencies---especially in the United States---is considerable. The Air Force was deeply involved with the U-2 Project, for example, even though it was technically under the jurisdiction of the CIA. The aerial photographs taken by U-2 missions were used to inform both military and civilian operations.} A military fight carries with it the implications of combat, aggression, and the possible prelude to an attack. A spy flight is instead a tool of civilian intelligence, whose purposes are often frustrated but never quite discouraged. The creation of the U-2 plane---a craft exclusively for civilian intelligence---is a pivotal moment in the history of espionage because the U-2 is the physical embodiment of the espionage norms it helped to define. Its \emph{raison d'\^etre} is an activity which is not permitted---or its altitude would be unnecessary---but its design emphasizes its peaceful reconnaissance purpose.

\section{The Open Skies policy}
From the beginning of the Cold War, aerial intelligence played a crucial role in American foreign policy. The United States considered aerial surveillance so important that it was willing to allow the Soviets to do it in return. President Eisenhower formally proposed a mutual inspection regime called Open Skies to the Soviet Union as a means of reducing security tensions. In his prepared remarks at the Geneva Summit of 1955, Eisenhower laid out the following plan:

	\begin{quote}
	\textbf{To give to each other a complete blueprint of our military establishments}, from beginning to end, from one end of our countries to the other, lay out the establishments and provide the blueprints to each other.
	\newline

	Next, to provide within our countries facilities for aerial photography to the other country---\textbf{we to provide you the facilities within our country, ample facilities for aerial reconnaissance}, where you can make all the pictures you choose and take them to your own country to study, you to provide exactly the same facilities for us and we to make these examinations, and by this step \textbf{to convince the world that we are providing as between ourselves against the possibility of great surprise attack}, thus lessening danger and relaxing tension. (emphasis mine)\footcite{eisenhower_president_1955}
	\end{quote}

The UN-sponsored Geneva Summit, formally the \emph{Conference on the Peaceful Uses of Atomic Energy}, was a comparatively hopeful period in the Cold War, a conference promoting scientific diplomacy where such a cooperative information exchange might have seemed feasible.\footcite[p.~27]{luscher_nuclear_2018} Nonetheless, the scale of this offer was astounding. Eisenhower offered to let the the Soviets install operational military bases in the United States from which they could surveil American national defense infrastructure, so long as they allowed the US to do the same.\footnote{Also included were reciprocal nuclear facility inspections, an immediate nonstarter in the USSR, but also controversial within the American administration back home. (\cite{prados_review_2015})} In effect, proposed a limited form of state-sanctioned espionage for explicitly defensive purposes---exactly the norm I am arguing exists in practice, albeit in a different form.

The offer was quickly rejected. In conversation over buffet that evening, Khrushchev politely countered that the plan would not further disarmament at all because it would merely confirm the fragmentary information that their respective intelligence services already possessed.\footcite{department_of_state_memorandum_1955} The response was telling---though it was in no way an endorsement of the activity, it both acknowledged that espionage was taking place between the two superpowers, and that it served a potentially de-escalatory purpose.

Khrushchev also claimed that the offer was primarily a propaganda move on the part of the US, but backed down when Eisenhower essentially dared him to call his bluff.\footcite[p.~2. This is a secondary source---the book cites a conversation by Eisenhower, quoted in NSC meeting memos that are stored in the Eisenhower library.]{bury_eisenhower_2014} The formal rejection came a few weeks later: ``While giving credit to this proposal's attempt to find a solution,'' USSR Council of Ministers chair Nikolai Bulganin wrote in his report to the Supreme Soviet, ``\textelp{} aerial photography cannot yield the expected results, since both our countries contain limitless expanses in which, if one desires, everything can be hidden.''\footcite{bulganin_meetings_1955} Though a lukewarm excuse, it was not a disdainful one. Earlier in the report, he had even said that the conference, as a whole, ``must be considered a definite success of the peace-loving forces.''

Was the offer genuine, or was it the stunt that Khrushchev accused it of being? In a book about the relationship between aerial reconnaissance and Eisenhower's New Look foreign policy, historian Helen Bury argues that a willingness to negotiate acceptable and enforceable agreements with foreign powers was central to his Cold War strategy.\footcite[p.~69]{bury_eisenhower_2014} More importantly, the intelligence that an Open Skies policy promised would have been invaluable in his ongoing effort to control defense spending, where Eisenhower was embattled by deeply paranoid American institutions.\footcite[p.~212]{bury_eisenhower_2014} His commitment to Open Skies in practice, however, is still a point of some historical debate. In reviewing her argument, intelligence historian John Prados sees a contradiction between her portrayal of Eisenhower's genuine desire for arms control and Eisenhower's role in significant nuclear buildup.\footcite[p.~233-234]{prados_review_2015} In the following chapter, I will present my own evidence that these two existed as part of the same strategy. Eisenhower used photo-reconnaissance---of the kind that an Open Skies policy would have legally permitted---to create an accurate assessment the military capabilities required to sufficiently deter a Soviet nuclear strike; as a result, the United States built far fewer missiles than it otherwise would have.

% Valentino: Actually, they were explicitly consistent. Ike thought that nukes were the cheapest way to counter the Soviet numerical advantage in conventional weapons (which were and are much more expensive). So Ike saw both as cost saving measures.

Regardless of Eisenhower's intent, it is clear that the United States would have far more to gain from such a policy than the USSR, and it made strategic sense for the Soviets to reject the offer. Writing about the history of CIA's imaging program, former imagery analyst David Lindgren notes that Khrushchev had the advantage of huge amounts of publicly-available information about US military installations through newspapers, maps, and official aerial photographs, while Soviet restrictions on press freedom denied symmetrical data.\footcite[p.~38]{lindgren_trust_2000} Allowing for an Open Skies policy would also have revealed that Soviet air defenses were not nearly as far along as their leadership claimed. Khrushchev's son would later write that his father's overriding concern at the time was to conceal Soviet weakness.\footcite[p.~133]{brugioni_eyes_2010}

Treaty or not, the United States was desperate for intelligence about the Soviet missile program, and aerial photo-reconnaissance was crucial to gaining it. The West's HUMINT capabilities within the Soviet Union were hamstrung by the intense surveillance under which their diplomats were placed, as well as the enormous size of the Russia itself.\footcite[p.~23]{lindgren_trust_2000} As the US increasingly relied on signals intelligence (SIGINT) to gather intelligence, the USSR began pursuing a more aggressive aerial defense policy, demonstrating their willingness to open fire on foreign reconnaissance aircraft. The increased cost associated with these missions did little to deter senior US policymakers from attempting to surveil the skies.\footcite[p.~4]{pedlow_central_1992}

% The Soviet Union was able to successfully destroy US Military aircraft during purported peacetime because in many cases those US aircraft were not supposed to be there in the first place. In many cases their flight path was a clear violation of Soviet territorial boundaries---a fact of which the administration was aware.\footcite{goodpaster_memorandum_1956}

In the rest of this chapter, I will argue that Eisenhower satisfied the need for intelligence that an Open Skies treaty would have provided by normalizing the collection of aerial intelligence and minimizing its diplomatic consequences. This process happened in two parts: first, through the routine resolution of military reconnaissance shootdowns, and second, by symbolically divorcing peacetime aerial espionage from military reconnaissance with the creation of a spy plane. For both the military and civilian missions, I will demonstrate how the even the most sensationalized incidents resulted in no more than ineffectual diplomatic protests, even though the spy plane could surveil locations that a ferret flight never could.

\section{Military reconnaissance flights}
\subsection{Timeline of early Cold War SIGINT missions}
Well before the U-2 Incident in 1960, the United States had been routinely flying military aircraft near Soviet airspace, and the USSR had been routinely shooting them down. For a conflict defined by its lack of direct skirmishes between the two superpowers, a surprising number of Americans died in these incidents, which are mostly forgotten to history. A Smithsonian investigation found that over the course of the Cold War more than 200 American pilots were lost to Soviet shootdowns---126 of which remain unaccounted for to this day.\footcite{glenshaw_secret_2017}

The purpose of these missions was to gather SIGINT identifying the location of critical radar installations along the Soviet border. Converted bombers---called ``ferrets'' by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were ``overflights,'' missions that deliberately violated Soviet airspace to collect intelligence at a much higher risk of failure. The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

Determining the quantity of early Cold War reconnaissance flights is difficult because they were a closely guarded secret at the time. Some flights, however, were inadvertently declassified---when they were shot down by enemy air defenses. Since my goal is to understand the diplomatic repercussions of these flights, an effective place to start is a list of reconnaissance flights that attracted diplomatic attention after being shot down. The original shootdown stories are available in the newspaper archives, and the incidents have been helpfully compiled by US Air Force Academy historian Dr. John Farquhar, who specializes in aerial reconnaissance and wrote his master's thesis on how it shaped early Cold War diplomacy. My choice of incidents to analyze in this chapter is based in part on his work.
% \newline

Additionally, the CIA history magazine \emph{Cryptologic Quarterly} analyzed the Soviet shootdowns of US reconnaissance aircraft in depth, creating a comprehensive (as far as we know) list of each American overflight that was shot down by the USSR (Table \ref{soviet-shootdowns}). The article was written in 1993, but not declassified until 2009. The author, Michael Peterson, uses two limiting criteria that perfectly suit our purposes: the only flights listed are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down. In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}

\begin{table}[ht]
\centering
\begin{tabular}{llr}
\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
1 July 1960       & USAF RB-47                    & Barents Sea                          \\
10 March 1964     & USAF RB-66                    & East Germany
\end{tabular}
\caption{Summary of Soviet Shootdowns, 1950-1964}
\label{soviet-shootdowns}
\end{table}

I cannot confirm whether Peterson's record is comprehensive. Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, more secret missions which for whatever reason the United States government has chosen to continue protecting to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete. First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is intended to be comprehensive. Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one. Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages. And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.

In the following section, I cross-reference the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact. A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function. Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which the shootdowns that involved espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.

\subsection{The Baltic Incident}
The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people. An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea. The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} Upon receiving the note, Secretary of State Dean Acheson remarked that Ambassador Vishinsky's manner was ``serious but not aggressive or antagonistic,'' and the Secretary ``recommended that publicity on our side should be avoided or, if unavoidable, minimized.''\footcite{kirk_ambassador_1950}

A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters. He blamed the Soviet Union directly, outlining the punishment that the Americans planned to impose:

\begin{quote}
	The United States Government demands that the the Soviet Government institute a prompt and thorough investigation of this matter in order that the facts set forth above my be confirmed to its satisfaction. The United States Government further demands that the most strict and categorical instructions be issued to the Soviet air force that there be no repetition, under whatever pretext, of incidents of this kind which are so clearly calculated to magnify the difficulties of maintaining peaceful and correct international relationships.

	The United States Government fully expects that, when its investigation is completed, the Soviet Government will express its regret for the unlawful and provocative behavior of its aviators, will see to it that those responsible for this action are promptly and severely punished and will, in accordance with established custom among peace-loving nations, pay appropriate indemnity for the unprovoked destruction of American lives and property.\footcite{the_associated_press_text_1950}
\end{quote}
This was a weak response---it imposed no consequences, threatened no sanctions, and demanded only the most \emph{pro forma} of reparations. In essence, the note demanded an apology and a promise not to repeat. There was absolutely no chance that the parties responsible on the Soviet side would be held accountable---in the most serious of cases, some lower-level officers might take the fall, but even that was unlikely to happen here. If the United States had followed through on its demand for reparations---the only material repercussion mentioned---then one could have argued that this statement had some teeth. Instead, the demands were rejected in totality by the USSR.\footcite{salisbury_kremlin_1950} As far as I can tell, no reparations were ever made and no further consequences were imposed as a result.

Both sides had reasons to make this a larger political issue. The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary. Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down (Figure \ref{baltic-flight-path}).\footcite[p.~7]{peterson_maybe_1993} At the time, some unnamed Navy and Air Force officers, speaking to reporters in response to the Soviet protest note, conceded that it was possible that a complete breakdown in the plane's navigation system might have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite[Technically, the  article credits the first statement about the possible electronic breakdown to ``Navy and Air Force officials'' and the 90 degrees statement to ``Aerial navigators.'' It is unclear whether the navigators in questions are with the US military, but it does not matter regardless.]{the_new_york_times_soviet_1950}

\begin{figure}
\centering
\includegraphics[scale=0.45]{baltic-flight-path.png}
\caption{USN PB4Y2 Flight Path}
\label{baltic-flight-path}
\end{figure}

Meanwhile, the United States knew that it hadn't lost a B-29, and that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down. Though the US was likely correct that the plane never flew over Latvia, it certainly got as close as it needed to in order to gather the necessary intelligence about Soviet military installations on the coast. No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady. Yet based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.

The American press had no such reservations. \emph{Washington Post} columnist Walter Lippmann wrote that it could not have been ``a local incident brought on by a local commander but [rather] that it was an act of Soviet policy. The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.''\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally. ``If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,'' Lippmann wrote, ``it would use a plane of a wholly different type.''\footcite{lippmann_baltic_1950}

A newspaper columnist is not a policymaker, but his argument was likely reflective of the political opinion that Americans held regarding espionage at the time: of course we do these things, we just do them \emph{better} than that---and if the US had been doing the thing that the Soviets claimed it was, then it would have been justified of them to open fire. ``Baltic Plane Mystery,'' a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders. ``Electronics make the old delimitations for border coastal flights ridiculous. A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.''\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further. ``The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system. And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.'' Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.

The most important consequence of the Baltic Incident was not diplomatic. As a direct result of the shootdown, President Truman, who until that point was largely uninformed about these missions, ordered a thirty-day halt to all reconnaissance flights. In that time, his office drafted an operating procedure that would balance the strategic need for reconnaissance with its potential diplomatic volatility.\footcite[p.~41]{farquhar_aerial_2015} The resulting memo from the Joint Chiefs of Staff created, with Truman's approval, the Special Electronic Airborne Search Project (SESP), which set clear delineations of responsibility between Navy and Air Force reconnaissance, and established executive responsibility for approving the missions.

Farquhar cites this as a landmark moment in the history of aerial reconnaissance. ``No longer would ferret operations be conducted ad hoc by the military services;'' he writes, ``from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S. foreign policy.''\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the ``serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.''\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: ``Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.''\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.

\subsection{Routine resolution}
Between the shootdown of the Navy Privateer in 1950 and the famous U-2 shootdown in 1960, the United States lost nine more aircraft to Soviet air defenses. Without analyzing each one at the same level of detail, I can demonstrate how the general attitudes displayed in response to the Baltic Incident became normalized. In each of the incidents listed, the pattern is the same---each side released dueling statements of protest, and the situation continued more or less as before.

Describe here are the next few peacetime reconnaissance shootdowns whose diplomatic consequences Farquhar analyzes. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is on Peterson's list as a ferret flight. Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}

In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan. The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane. Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: ``Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.''\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets. The move demonstrated Eisenhower's ability to resist the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015}

Without belaboring the point, suffice to say that Farquhar concludes that after this moment in 1954, the resolution of these incidents became less tense. ``International incidents posed by shoot downs of reconnaissance aircraft still acted as a barrier in the path of d\'etente; but, by late 1954, overriding strategic concerns dictated a move toward breaking the cycle of hostility.''\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.

One incident in the period between 1950-1954 does not fit the broader pattern. On November 20, 1951, an American C-47 transport was forced down in Hungary. This incident is omitted from Peterson's list because, despite Soviet cries of ``spies and saboteurs,'' there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment. While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.\footcite{the_united_press_soviet_1951}

I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure. The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country. Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid `selected persons residing in Soviet bloc states or refugees who wanted to form armed units' in opposition to Communism.''\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released. The contrast between this response and the one from just a year prior, when the US demanded ``a prompt and thorough investigation'' of the Baltic Incident, is stark.

We also know that the United States conducted many more overflights than the failed ones that are listed here, using a variety of retrofitted planes. In 2001, the Air Force held a symposium honoring and recollecting the pilots who flew the overflights on Soviet territory in the early Cold War, an effort in which they were joined by the British in 1954.\footcite[p.~v]{hall_early_2003} Many of the presenters spoke about entire reconnaissance projects that apparently didn't even register diplomatically. Project Heart Throb, which outfitted RB-57s with surveillance equipment, flew between 15 and 19 missions over Eastern Europe from 1955 to 1956, in the recollection of Gerald E. Cooke, a Air Force pilot assigned to the project.\footcite[p.~194]{hall_early_2003} Another project from around the same time, Slick Chick, set up F-100 jets with rapid-fire 20mm cameras. The pilot who presented this project at the symposium, Cecil Rigby, personally flew two such missions and estimated that there were six Slick Chick missions total.\footcite[p.~176]{hall_early_2003}

The USSR knew about these missions, which were tracked by radar and often chased by MiG fighters. Unlike some of the flights that were shot down, these overflights were clear violation of their territorial sovereignty. We will likely never know exactly how many such missions were flown. The Truman and Eisenhower administrations were shockingly unsuccessful at recovering the pilots who were held as POWs by the Soviets. Dino Brugioni, a senior CIA officer involved in the aerial imagery program, paints a grim picture of the institutional accountability involved when the flights went bad. In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots ``was broached only perfunctorily in relation to other things being discussed.''\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation. ``And because these were secret missions conducted by a field command,'' Brugioni writes, ``a change of field commander meant that the fate of the men was soon forgotten.''\footcite[p.~72]{brugioni_eyes_2010} With missions so secret that we abandoned the men who flew them, we can have no expectation of finding a complete record today.

Even without a complete list of flights, it is clear that the territorial boundaries of the USSR were routinely violated to generate critical intelligence for NATO powers in an early, relatively volatile stage of the Cold War. As time passed these flights appear to have become more accepted, not less. Diplomats of both nations were, as a matter of policy, able to paper over these issues in service of advancing other interests. When the shootdown caused unavoidable diplomatic consequences---such as when it was well-publicized---the countries exchanged the necessary diplomatic cables and engaged in the requisite saber rattling. These incidents undoubtedly altered the climate of Cold War diplomacy, but the fact remains that in 10 years of reconnaissance, the only action the Soviets took to keep the United States from conducting these incredibly provocative reconnaissance flights was to shoot them down.

% Valentino: Can you talk about other ways the Sovierts might have responded. What other kinds of actions might have seemed reasonable under different norms? Maybe give an example of a Soviet response to something apparently similar on its face, but clearly out of the espionage domain, so the response was much harsher.

\section{Civilian spy flights}
\subsection{Project Aquatone}
While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow. What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict. What they needed was a spy plane.

The key feature of a spy plane is that it is designed to be minimally provocative. Any plane can be retrofitted with cameras or radios that allow it to perform useful reconnaissance functions---what separates a spy plane is that there can can be no doubt (in the mind of the target) that it is being used for anything \emph{but} reconnaissance. For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response. To build such a aircraft, Eisenhower approved Project Aquatone, the code name for the creation of the U-2.

From its conception, the U-2 was purpose-built to be read as a civilian aircraft. Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force. ``I want this whole thing to be a civilian operation,'' Eisenhower wrote, when settling an operational dispute between the two departments. ``If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.''\footcite[p.~60. The original source for this quote is an \emph{OSA History} that requires codeword clearance. It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to ``truthfully deny,'' in the phrasing of the CIA, that any US ``military planes'' had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}

Having the CIA run the program---ostensibly for the purpose of deception---changed the nature of the program itself. If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient. The ``weather reconnaissance'' excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold. ``It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft'' read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important. The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.

The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about ``the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.''\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called ``missile gap,'' concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a ``bomber gap,'' which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960. Nonetheless, Eisenhower ordered two more overflights in 1960, one on April \nth{9} and the other on May \nth{1}. Only one of them returned.

% It turned out however, that the Soviets opted to lodge their diplomatic protests privately, and soon they stopped protesting them at all.\footcite[p.~42]{lindgren_trust_2000}

\subsection{The U-2 Incident}
On the morning of May 1, 1960, Soviet Air Defense forces detected a high-altitude aircraft flying over Soviet Tajikistan. Though they were were not yet able to conclusively identify it, the foreign aircraft was an American U-2 spy plane performing an reconnaissance overflight of the USSR---at a time when the Soviets were deeply embarrassed by their inability to prevent such flights.\footcite{orlov_u-2_2007} The previous U-2 overflight on April 9, less than a month prior, had lasted a full six hours and the aftermath caused an upheaval in the Soviet military. Leadership ordered an investigative commission to root out the shortcomings of the Air Force and Air Defense agencies, and a series of charts were drawn up anticipating the routes of future U-2 flights. The next time one came through, the Soviets were absolutely determined to shoot it down.

\begin{figure}
\centering
\includegraphics[scale=0.3]{powers-flight-path.jpg}
\caption{Operation Grand Slam Flight Path}
\label{powers-flight-path}
\end{figure}

The Soviets still proved unable to immediately stop the flight on May \nth{1}. A missile battalion in the plane's path was not on alert duty. Armed fighter aircraft were in the wrong positions. A frantic attempt to have another fighter plane literally ram it out of the sky was scrapped when the pilot failed to make visual contact.\footcite{orlov_u-2_2007} The highest possible levels of Soviet leadership were actively involved with the mission as it was taking place. A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft on the day of a Soviet national holiday, and just two weeks before a summit conference in Paris, as a political provocation.'' \footcite{orlov_u-2_2007}
